Spring haul-out behavior of seals in the Bering and Chukchi Seas: implications for abundance estimation

### 2024-09-03 analysis/paper/London_HauloutBehavior.Rmd
remove british spelling of behavior from title

title: Spring haul-out behaviour of seals in the Bering and Chukchi Seas: implications for abundance estimation

@@

  activities occur in spring (April -- June) as sea ice begins to melt and

@@

  barbatus_), ribbon (_Histriophoca fasciata_), and spotted seals (_Phoca
  largha_) in the Bering and Chukchi seas. Because ringed seals (_Phoca (pusa)
  hispida_) can be out of the water but hidden from view in snow lairs analysis
  of their 'availability' to surveys requires special consideration;
  therefore,they were not included in this analysis. Using generalized linear
  mixed pseudo-models to properly account for temporal autocorrelation, we fit
  models with covariates of interest (e.g., day-of-year, solar hour, age-sex
  class, wind speed, barometric pressure, temperature, precipitation) to examine
  their ability to explain variation in haul-out probability. We found evidence
  for strong diel and within-season patterns in haul-out behavior, as well as
  strong weather effects (particularly wind and temperature). In general, seals
  were more likely to haul out on ice in the middle of the day and when wind
  speed was low and temperatures were higher. Haul-out probability increased
  through March and April, peaking in May and early June before declining again.
  The timing and frequency of haul-out events also varied based on species and
  age-sex class. For ribbon and spotted seals, models with year effects were
  highly supported, indicating that the timing and magnitude of haul-out
  behavior varied among years. However, we did not find broad evidence that
  haul-out timing was linked to annual sea-ice extent. Our analysis emphasizes
  the importance of accounting for seasonal and temporal variation in haul-out
  behavior, as well as associated environmental covariates, when interpreting
  the number of seals counted in aerial surveys.

>>>>

title: "Spring haul-out behavior of seals in the Bering and Chukchi Seas: implications for abundance estimation"

@@

  activities occur in spring (April through June) as sea ice begins to melt and

@@

  barbatus_), ribbon (_Histriophoca fasciata_), and spotted seals (_Phoca largha_) in
  the Bering and Chukchi seas. Because ringed seals (_Phoca (pusa) hispida_) can
  be out of the water but hidden from view in snow lairs analysis of their
  'availability' to surveys requires special consideration; therefore, they were
  not included in this analysis. Using generalized linear mixed pseudo-models to
  properly account for temporal autocorrelation, we fit models with covariates
  of interest (e.g., day-of-year, solar hour, age-sex class, wind speed,
  barometric pressure, temperature, precipitation) to examine their ability to
  explain variation in haul-out probability. We found evidence for strong diel
  and within-season patterns in haul-out behavior, as well as strong weather
  effects (particularly wind and temperature). In general, seals were more
  likely to haul out on ice in the middle of the day and when wind speed was low
  and temperatures were higher. Haul-out probability increased through March and
  April, peaking in May and early June before declining again. The timing and
  frequency of haul-out events also varied based on species and age-sex class.
  For ribbon and spotted seals, models with year effects were highly supported,
  indicating that the timing and magnitude of haul-out behavior varied among
  years. However, we did not find broad evidence that haul-out timing was linked
  to annual sea-ice extent. Our analysis emphasizes the importance of accounting
  for seasonal and temporal variation in haul-out behavior, as well as
  associated environmental covariates, when interpreting the number of seals
  counted in aerial surveys.
